### 2019-09-02 Manuscripts/post-cogpsych/post-cogpsych.tex
Updated draft Exp 3 with correlations.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Contrast from other
known categories should play a role. Indeed, scientists in other fields, such
as marketing \citep{berger16} and sociology \citep{rogers03}, highlight the
critical importance of contrast in the creation of new ideas.

@@

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another.

@@

Although people are {\em capable} of creating 
new categories, it is not entirely clear whether (and if so, how) new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over stimulus space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast, which is precisely the factor of study in this article.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

@@

As noted above, the constraint that new concepts should differ from what is
already known has been largely overlooked in previous work. This is no doubt in
part due to the vague definition of what it means for a concept to be
``different'': A generated category may be different from what is already known
in any number of respects. Towards providing a more precise definition of the
role of contrast in generation, we formalized contrast in a novel exemplar
model, PACKER ({\em Producing Alike and Contrasting Knowledge using Exemplar
  Representations}). PACKER explains category generation as a balance between
two fundamental constraints: The category to be generated should not be similar
to known categories, and exemplars within each category should be similar to one
another. These ideas are implemented within the well-studied exemplar framework
-- the PACKER model is an extension of the influential Generalized Context Model
of categorization \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

@@

\begin{equation} s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k
\left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and
$\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in
similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling
the spread of exemplar generalization. For simplicity, attention will be
distributed uniformly in our simulations (unless otherwise noted). The value of
$r$ depends on the nature of the experimental conditions being simulated: $r=1$
is appropriate for separable dimensions, whereas $r=2$ is appropriate for
integral dimensions \citep[e.g.,][]{shepard1964attention,garner1974processing}.
In our simulations, we set $r=1$ due to the separable nature of the stimulus
dimensions used in our experiments (see Figure \ref{fig:e1-conditions}).

@@

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}
% 
%where $\theta$ ($\theta \geq 0$) is a free parameter controlling response
%determinism.
It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, as will be discussed in more detail below, PACKER is formally
equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic}
when $\theta_c = 0$ and $\theta_t > 0$. Likewise, when
$\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a
contrast-only generation mode, relying exclusively on contrast when generating
new categories. When $f(x_j) < 0$ for all $x_j$ (regardless of class
membership), a ``pure-packing'' approach is yielded, generating items in
unoccupied areas of the domain. Thus, the proposed framework may be used to
describe a wide variety of qualitatively distinct generation strategies.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
%: Both models are
%exemplar-based, and both models generate new items according to their similarity
%to known members of the target class. However, PACKER diverges from the
%copy-and-tweak model by including a contrast mechanism, enabling generation
%according to dissimilarity to members of opposing categories. As a consequence,
%copy-and-tweak can be realized as a parameterization of the PACKER model that is
%insensitive to category contrast. Specifically, when $\theta_c = 0$ and
%$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
%$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
%items, and is mathematically equivalent to a copy-and-tweak approach.

\label{section:copytweak-packer} The PACKER model is similar to the copy-and-tweak model reported by \cite{jern2013probabilistic}.
An advantage of formulating copy-and-tweak as a special case parameterization of PACKER. In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and thus, we
leave extending PACKER to incorporate distributional factors for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories. We
adopt the formalization given by \cite{tenenbaum01}, where the
representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is the
relative amount of evidence that is provided by $y$ for a given hypothesis $h$
in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in \mathcal{H}^c}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

In our simulations, we set the prior mean $\mu_0$ to the center of the stimulus
space and the prior variance to be isotropic ($\Sigma_0 = \lambda \mathbf{I}$ where
$\lambda$ is a free parameter and $\mathbf{I}$ is a $d$-by-$d$ identity matrix
with $d$ representing the number of dimensions or features in the domain). The
$\kappa$ and $\nu$ parameters are freely estimated within the constraints
$\kappa > d-1$ and $\nu > 0$.

@@

We believe that the failure to replicate dimension correlation emulation is
because participants in \cite{jern2013probabilistic} could differentiate the
generated category on a third dimension (hue) to maintain the statistical
regularities on the other two dimensions. In addition, in our XOR condition the
stimuli were constrained to the corners of the feature space, whereas in the
positive diagonal condition of \cite{jern2013probabilistic} the stimuli had no
such constraint. Although the correlation in the XOR condition is significantly
negative, it is clear from the box-plot in Figure~\ref{fig:e1-statsboxes} that
it would be inappropriate to make a strong conclusion (e.g., the median is close
to zero). However, we can conclude with confidence that there are situations
where people do not emulate the distributional structure of the given category.
This indicates that there is more to category generation than the emulation of
distributional structure of other categories in the domain. Further, as we will
discuss in more detail in the model-based analysis section, this is expected by
our proposal that contrast is a fundamental principle in categorization.

@@

To test whether category contrast influences the distributional structure of
generated categories, we sought to identify conditions in which differences in
the distributional structure of generated categories cannot be explained by the
distributional structure of the experimenter-defined category. We created two
new category types (depicted in Figure \ref{fig:e2-conditions}) that possess
identical distributional structures (both are tight clusters of examples with no
correlation between features), as they only differ in their Y-axis position: the
`Bottom' category lies near the bottom of the space, and the `Middle' category
lies in the center. The distributional equality of these conditions is key to
the design of the experiment: If the distributional structure of previously
learned categories were the only influence on the generated categories, we
should observe no difference in the categories participants generate between
these two conditions. Will participants distribute their generated category
differently between conditions due to the differences in the available empty
stimulus space for generating a new category?

If category contrast influences the distributional structure of the categories
people generate, then we should observe different types of categories according
to the shape of the space that is {\em unoccupied} by members of previously
learned categories. The difference in the Y-axis position between the Bottom and
Middle conditions produces a considerable change to the shape of the unoccupied
space. Participants assigned to learn the Bottom category should be less likely
to generate exemplars into the lower regions of the stimulus space (as these
areas possess greater similarity to members of the Bottom category), preferring
instead to distribute exemplars across the upper region of the space. This
constraint is lifted in the Middle condition, as the Middle category exemplars
are equidistant to the upper and lower regions of the space. Accordingly,
participants should be more likely to utilize both of these areas. Thus, if
category contrast influences the distributional structure of generated
categories, we should observe more participants in the Middle condition that
generate examples above {\em and} below the experimenter-defined category.

@@

%Contrast model distinction
\section{Experiment 3: Differentiating Contrast}
\label{section:diffcon}
Experiments 1 and 2 clearly established the importance of contrast in category
generation. Follow-up model-based analyses illustrated that both contrast models account for participant performance better than models that do not take into
account contrast. We also found more support for the hypothesis of contrast as
representativeness versus contrast as exemplar dissimilarity across the
different Alpha conditions.

While the representativeness model fits better than PACKER, and in spite of
their fundamentally different structures, both contrast models make qualitatively similar predictions for Experiments 1 and 2. To illustrate this, Figure \ref{fig:firstbetas} shows how the modal generating probabilities for the first Beta exemplar are similar across both contrast models.\footnote{Model parameters are set to the optimized values from Table \ref{table:global-model-fits}.} The highest density regions for both models appear to be near the corners and edges. This is particularly the case for the representativeness model, which almost exclusively predicts generation at the corners, while PACKER can generate the first novel exemplars toward the inner regions of the stimulus space in the XOR condition.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=\textwidth]{figs/firstbetas.pdf}
      \caption{Contrast model predictions indicating probability of generating
        the first Beta exemplar for each Alpha condition. Darker shades of
        blue indicate higher probabilities (normalized within each plot.)}
      \label{fig:firstbetas}
    \end{center}
\end{figure}

These similarities are perhaps unsurprising -- each of the modal generation
regions are maximally distant from the Alpha exemplars. However, recall that
unlike PACKER, the representativeness model does not explicitly use a measure sensitive to distance from the location of exemplars in the contrasting category. The general pattern of producing exemplars dissimilar from the contrast category is because a unimodal category structure is assumed by the model and it is not strictly constrained by the distances to exemplars of the contrasting categories. Consequently, it is possible to design a configuration of Alpha exemplars where the representativeness model would predict the generation of novel exemplars
that are \emph{more similar} to the learned exemplars.

In this section, we focus on one category type -- the Corner
category -- to distinguish between the two competing accounts of contrast in
category generation. Here, each Alpha exemplar is located at the extreme corners of the feature space (see Figure \ref{fig:e3-squares}a). We first seek to identify what predictions each contrast model generates under this condition,
followed by an category generation task, and a model-based analysis of the two contrast models given the behavioral data.

In the Corner condition, the representativeness model assumes an underlying
Alpha category distribution that is centered on the feature space, with
probability densities highest in the center and lowest in the corners. Since the representativeness of novel Beta exemplars generally increase when the
likelihood of the underlying Alpha category distribution decreases, the regions
of highest Beta representativeness are consequently located at the corners where
these Alpha likelihoods are lowest (Figure \ref{fig:e3-squares}b.)

Conversely, PACKER predicts a completely different result. With $\theta_c > 0$,
the probability of generating a Beta exemplar increases with the dissimilarity
from all Alpha exemplars. This results in highest Beta generation probabilities
in the inner regions of the feature space, where the candidate exemplars are
maximally different from the Alpha exemplars (Figure \ref{fig:e3-squares}c.)

@@

Another factor we examine with this category structure is the effect of
boundaries. The representativeness model prefers to generate categories located
at the boundaries of the stimulus features. If at least one of these boundaries
is removed by utilizing a toroidal feature (one with no boundaries), such as the
rotation of a line.[Note to Xian: Something weird going on here, remember to edit.] The representativeness model will strongly prefer to
generate categories on the non-toroidal or bounded feature as it still has a
boundary. Experiment 3 manipulates the stimulus space. One condition uses the
same stimuli as before, squares varying in lightness and size, both of which are
bounded features. The other condition use a new set of stimuli: circles that
vary in size (a bounded feature) and orientation (an unbounded feature; Figure
\ref{fig:e3-circles}a). Before describing the details of the experiment, we
examine each model's predictions with stimuli that comprise one bounded and one
unbounded feature.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=.8\textwidth]{figs/e3-circles.png}
      \caption{Circle stimulus domain (a) and predictions from contrast models
        PACKER (b) and the representativeness model (c). The unbounded feature
        (orientation) is represented on the y-axis, while the bounded feature
        (size) is represented on a x-axis. As with the square stimuli, feature
        and direction assignment were counterbalanced across participants (in
        the displayed example, circles are rotated clockwise up the y-axis.) }
      \label{fig:e3-circles}
    \end{center}
\end{figure}

Compared to stimuli with two bounded features, PACKER makes a similar prediction
in generating the first Beta exemplar when one feature is unbounded, with the
curious exception that the highest density regions appear compressed along the
y-axis (for simplicity, we always assign the y-axis as the toroidal feature in
the text and figures). Due to the unbounded nature of orientation, the maximum
distance from the Alpha exemplars is half of the full range of the feature --
for an Alpha exemplar oriented at 0 radians, the furthest another exemplar can
be situated is $\pi$ radians within the full range of orientation is $2\pi$
radians.

For the representativeness model, adding an unbounded feature changes the
probability distribution for the underlying Bayesian model. Whereas it is still
Gaussian in one feature, the other feature is toroidal, meaning that it wraps
back to $0$ radians once it reaches $2\pi$ (we have conducted a change of
variables where $y \rightarrow y \mod 2\pi $, where $\mod$ is the modulus
operator). As we cross $2\pi$ and reach $0$ again, the remaining probability
density also gets wrapped around, meaning that the probability density assigned
by the Gaussian distribution on the toroidal dimension at $2\pi$ is actually
assigned to $0$ (and $2\pi + \epsilon$ to $\epsilon$, and so on). This is called the
{\em wrapped} Gaussian distribution \citep{mardia1972}. Note that the often used
distribution for circular dimensions, the Von Mises distribution, is an
approximation to the wrapped Gaussian distribution. See Appendix B for
implementational details of the representativeness model in toroidal space.

>>>>

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.\footnote{It
  is worth noting that sensitivity to feature correlations is not universal and
  can be dependent on factors such as the salience of correlations and task
  demands \citep[e.g., see][]{chin2002effect,malt1984correlated,murphy2004big}.}
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. These statistical regularities
reflect general structural patterns across old and new categories and are not
restricted to strictly quantitatively similar properties.

Although statistical regularity is an important characteristic of generating new
categories, it cannot be the only one. Taken to the extreme, the best ``new''
category in terms of having the same statistical regularities to other
categories would be identical to a known category that is representative of the
domain (and thus, not new at all). Contrast from other known categories should
play a role. Indeed, scientists in other fields, such as marketing
\citep{berger16} and sociology \citep{rogers03}, highlight the critical
importance of contrast in the creation of new ideas.

@@

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another.

@@

Although people are {\em capable} of creating 
new categories, it is not entirely clear whether (and if so, how) new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over feature space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast, which is precisely the factor of study in this article.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

@@

As noted above, the constraint that new concepts should differ from what is
already known has been largely overlooked in previous work. This is no doubt in
part due to the vague definition of what it means for a concept to be
``different'': A generated category may be different from what is already known
in any number of respects. Towards providing a more precise definition of the
role of contrast in generation, we formalized contrast in a novel exemplar
model, PACKER ({\em Producing Alike and Contrasting Knowledge using Exemplar
  Representations}). PACKER explains category generation as a balance between
two fundamental constraints: The exemplars of the category to be generated
should not be similar to known categories, and exemplars within each category
should be similar to one another. These ideas are implemented within the
well-studied exemplar framework -- the PACKER model is an extension of the
influential Generalized Context Model of categorization
\citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

@@

\begin{equation} s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k
\left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and
$\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in
similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling
the spread of exemplar generalization. For simplicity, in our simulations
attention will be distributed across each dimension uniformly (unless otherwise
noted). The value of $r$ depends on the nature of the experimental conditions
being simulated: $r=1$ is appropriate for separable dimensions, whereas $r=2$ is
appropriate for integral dimensions
\citep[e.g.,][]{shepard1964attention,garner1974processing}. In our simulations,
we set $r=1$ due to the separable nature of the stimulus dimensions used in our
experiments (see Figure \ref{fig:e1-conditions}).

@@

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}
% 
%where $\theta$ ($\theta \geq 0$) is a free parameter controlling response
%determinism.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
\label{section:copytweak-packer}
%: Both models are
%exemplar-based, and both models generate new items according to their similarity
%to known members of the target class. However, PACKER diverges from the
%copy-and-tweak model by including a contrast mechanism, enabling generation
%according to dissimilarity to members of opposing categories. As a consequence,
%copy-and-tweak can be realized as a parameterization of the PACKER model that is
%insensitive to category contrast. Specifically, when $\theta_c = 0$ and
%$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
%$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
%items, and is mathematically equivalent to a copy-and-tweak approach.

It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, PACKER is formally equivalent to the copy-and-tweak model proposed by
\cite{jern2013probabilistic} when $\theta_c = 0$ and $\theta_t > 0$ (in fact, it
becomes a continuous-dimension adaptation of the original model). Likewise,
when $\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a contrast-only
generation mode, relying exclusively on contrast when generating new categories.
When $f(x_j) < 0$ for all $x_j$ (regardless of class membership), a
``pure-packing'' approach is yielded, generating items in unoccupied areas of
the domain. Thus, the proposed framework may be used to describe a wide variety
of qualitatively distinct generation strategies.

By formalizing a model family where PACKER and copy-and-tweak are different
parameterizations of models within the same framework, the comparison between
PACKER and copy-and-tweak provides a test of the explanatory value of the
contrast mechanism based on exemplar dissimilarity: The account provided by
copy-and-tweak will only equal that of PACKER if the contrast mechanism does not
offer an advantage (i.e., if $\theta_c > 0$ significantly improves model fits).
Note that the purpose of the article is to explore and formally analyze the role
of contrast in categorization and thus, we leave extending PACKER to incorporate
distributional factors for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories.
The general idea of representativeness is not new in categorization. For
instance, the literature on the \emph{graded structure} of categories explores
how some exemplars are better representations of their categories
\citep{palmeri2001central,barsalou1985ideals}. In this article we adopt the
specific formalization of representativeness given by \cite{tenenbaum01}, where
the representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is
the relative amount of evidence that is provided by $y$ for a given hypothesis
$h$ in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in \mathcal{H}^c}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

In our simulations, we set the prior mean $\mu_0$ to the center of the feature
space and the prior variance to be isotropic ($\Sigma_0 = \lambda \mathbf{I}$ where
$\lambda$ is a free parameter and $\mathbf{I}$ is a $d$-by-$d$ identity matrix
with $d$ representing the number of dimensions or features in the domain). The
$\kappa$ and $\nu$ parameters are freely estimated within the constraints
$\kappa > d-1$ and $\nu > 0$.

@@

We believe that the failure to replicate the emulation of dimension correlation is
because participants in \cite{jern2013probabilistic} could differentiate the
generated category on a third dimension (hue) to maintain the statistical
regularities on the other two dimensions. In addition, in our XOR condition the
stimuli were constrained to the corners of the feature space, whereas in the
positive diagonal condition of \cite{jern2013probabilistic} the stimuli had no
such constraint. Although the correlation in the XOR condition is significantly
negative, it is clear from the box-plot in Figure~\ref{fig:e1-statsboxes} that
it would be inappropriate to make a strong conclusion (e.g., the median is close
to zero). However, we can conclude with confidence that there are situations
where people do not emulate the distributional structure of the given category.
This indicates that there is more to category generation than the emulation of
distributional structure of other categories in the domain. Further, as we will
discuss in more detail in the model-based analysis section, this is expected by
our proposal that contrast is a fundamental principle in categorization.

@@

To test whether category contrast influences the distributional structure of
generated categories, we sought to identify conditions in which differences in
the distributional structure of generated categories cannot be explained by the
distributional structure of the experimenter-defined category. We created two
new category types (depicted in Figure \ref{fig:e2-conditions}) that possess
identical distributional structures (both are tight clusters of examples with no
correlation between features), as they only differ in their Y-axis position: the
`Bottom' category lies near the bottom of the space, and the `Middle' category
lies in the center. The distributional equality of these conditions is key to
the design of the experiment: If the distributional structure of previously
learned categories were the only influence on the generated categories, we
should observe no difference in the categories participants generate between
these two conditions. Will participants distribute their generated category
differently between conditions due to the differences in the available empty
feature space for generating a new category?

If category contrast influences the distributional structure of the categories
people generate, then we should observe different types of categories according
to the shape of the space that is {\em unoccupied} by members of previously
learned categories. The difference in the Y-axis position between the Bottom and
Middle conditions produces a considerable change to the shape of the unoccupied
space. Participants assigned to learn the Bottom category should be less likely
to generate exemplars into the lower regions of the feature space (as these
areas possess greater similarity to members of the Bottom category), preferring
instead to distribute exemplars across the upper region of the space. This
constraint is lifted in the Middle condition, as the Middle category exemplars
are equidistant to the upper and lower regions of the space. Accordingly,
participants should be more likely to utilize both of these areas. Thus, if
category contrast influences the distributional structure of generated
categories, we should observe more participants in the Middle condition that
generate examples above {\em and} below the experimenter-defined category.

@@

%Contrast model distinction
\section{Experiment 3: Contrasting Contrast}
\label{section:diffcon}
Experiments 1 and 2 clearly established the importance of contrast in category
generation. Follow-up model-based analyses illustrated that both contrast models account for participant performance better than models that do not take into
account contrast. We also found more support for the hypothesis of contrast as
representativeness versus contrast as exemplar dissimilarity across the
different Alpha conditions.

While the representativeness model fits better than PACKER, and in spite of their fundamentally different structures,
both contrast models make qualitatively similar predictions for Experiments 1 and 2. To illustrate this, Figure
\ref{fig:firstbetas} shows how the models' generating probabilities for the first Beta exemplar are similar across both
contrast models.\footnote{Model parameters are set to the optimized values from Table \ref{table:global-model-fits}.}
Across the five Alpha conditions of Experiments 1 and 2, the first Beta exemplars according to the contrast models were
generally more probable near the corners and edges. This is particularly the case for the representativeness model,
which almost exclusively predicts generation at the corners, while PACKER can predict generation in the center of the
feature space in the XOR condition.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=\textwidth]{figs/firstbetas.pdf}
      \caption{Contrast model predictions indicating probability of generating
        the first Beta exemplar for each Alpha condition. Darker shades of
        blue indicate higher probabilities (normalized within each plot).}
      \label{fig:firstbetas}
    \end{center}
\end{figure}

The effect of contrast for both models is similar in that they both tend to produce exemplars that are distant from the
Alpha category. However, this is realized in different ways for the two models. The representativeness model does not
explicitly use a measure sensitive to distance from the location of exemplars in the contrasting category. Instead, it
is the assumption of unimodal distributions that results in the model producing novel exemplars dissimilar to learned
exemplars. Consequently, it is possible to design a configuration of Alpha exemplars where the representativeness model
predicts novel exemplars should be \emph{similar} to the learned exemplars. However, PACKER cannot predict this because
it assumes that the similarity weight for contrasting exemplars is negative.

In this section, we focus on a different category type -- the Corner category -- that contrasts the two models of
contrast. Here, each Alpha exemplar is located at the corners of the feature space (see Figure \ref{fig:e3-squares}a).

We first identify the predictions of each model in this condition. Second, we test these contrasting predictions in a
behavioral experiment. We then conduct a model-based analysis of the two contrast models.

In the Corner condition, the representativeness model estimates the underlying Alpha category distribution to be
centered on the feature space, with probability densities highest in the center and lowest in the corners. The
representativeness of novel Beta exemplars generally decreases with the likelihood of the underlying Alpha category
distribution. Despite the Alpha exemplars being in the corners, the predicted mean will be near the center
of the Alpha category (the center of the space; See \ref{fig:e3-squares}b). As such, the Alpha likelihood is lowest in
the corners, and thus, the representativeness of the Beta category will be largest in the corners.

Conversely, PACKER predicts a completely different result. With PACKER, the probability of generating a Beta exemplar
increases with the dissimilarity from all Alpha exemplars. This results in Betas generated in the center of feature
space, where the candidate exemplars are as far as possible from the Alpha exemplars (Figure \ref{fig:e3-squares}c).

@@

Another aspect of the contrast in the models' predictions is the role of the boundaries of feature space. The
representativeness model prefers to generate categories at the boundaries of feature space. PACKER is less sensitive to
boundaries. Boundaries tend to be where distance to the contrasting category exemplars is maximized. But, as in cases
like the Corner condition (Figure \ref{fig:e3-squares}a), this is not always the case; Beta exemplars should be as far
from the Alphas as possible, which is maximized in the center of feature space.

Thus, in this experiment, we also examine how making a boundary toroidal (e.g., line rotation) affects the prediction of
each model. The representativeness model prefers to generate categories on the feature that still has a boundary.
PACKER, in contrast, should continue to generate categories that are most distant from the learn category regardless of
the presence or absence of feature boundaries -- in this condition it would be the center of the feature space.
Experiment 3 manipulates whether one feature is bounded or toroidal (always displayed as the Y-axis). One condition uses
the same stimuli as before, squares varying in lightness and size, both of which are bounded features. The other
condition use a new set of stimuli: circles that vary in size (a bounded feature) and orientation (an unbounded feature;
Figure \ref{fig:e3-circles}a). Before describing the details of the experiment, we explain in detail each model's
predictions given the Corner Alphas for a feature space where one feature is bounded and the other is toroidal.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=.8\textwidth]{figs/e3-circles.pdf}
      \caption{Circle stimulus domain (a) and predictions from contrast models
        PACKER (b) and the representativeness model (c). The unbounded feature
        (orientation) is represented on the y-axis, while the bounded feature
        (size) is represented on a x-axis. As with the square stimuli, feature
        and direction assignment were counterbalanced across participants (in
        the displayed example, circles are rotated clockwise up the y-axis). }
      \label{fig:e3-circles}
    \end{center}
\end{figure}

Compared to stimuli with two bounded features, PACKER makes a similar prediction in generating the first Beta exemplar
when one feature is unbounded, with the curious exception that the highest density regions appear compressed along the
toroidal feature. Due to the toroidal nature of orientation, the maximum distance from the Alpha exemplars is half of
the full range of the feature -- for an Alpha exemplar oriented at 0 radians, the furthest distance to another exemplar
is $\pi$ radians.

For the representativeness model, adding a toroidal feature changes the probability distribution for the underlying
Bayesian model. For the toroidal feature, space wraps back to $0$ radians once it reaches $2\pi$ (we have conducted a
change of variables where $y \rightarrow y$ mod $2\pi$, where mod is the modulus operator). As we cross $2\pi$ and reach
$0$ again, the remaining probability density also gets wrapped around, meaning that the density at $2\pi$ is added to
$0$ (and $2\pi + \epsilon$ to $\epsilon$, and so on). This is called the {\em wrapped} Gaussian distribution
\citep{mardia1972}. Note that the more familiar distribution for circular features, the Von Mises distribution, is an
approximation to the wrapped Gaussian distribution. See Appendix B for implementational details of the
representativeness model in a bounded and toroidal feature.

The consequence of employing a wrapped Gaussian is presented in Figure \ref{fig:e3-circles}c. Edge effects along the
unbounded feature (displayed as the vertical axis) are no longer present since the edges themselves no longer exist.
Instead, the probability of generating the first Beta exemplar is bimodally distributed at the edges of the bounded
feature (displayed as the horizontal axis). Thus, the representativeness model and PACKER make orthogonal predictions
for how a toroidal dimension should affect category generation.
